Follow-up Samples

### 2017-02-12 submission/manuscript_outline_20161024.Rmd
incorporate updated SRA and corrected RF runs into manuscript

**Results:** The bacterial microbiome in the pre and post surgery samples of the individuals with adenoma were similar while those with carcinoma were not (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).  There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  The post surgery sample for both models had a decrease in the positive probability for either lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`).   
**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.  With individuals with carcinoma having more drastic changes to the overall community then those with adenoma.  However, changes to commensal bacteria were common to individuals with either adenoma or carcinoma suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

***Common OTUs to both Models***
The reduced models were built based on the most important variables to either classification model.  For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].  There were a total of `r length(common_vars_summary$otu)` OTUs that were common to both models.  Of these OTUs the most common taxonomic identifications were to  `r common_vars_summary['Otu000012', 'lowest_ID']`, `r common_vars_summary['Otu000005', 'lowest_ID']`, `r common_vars_summary['Otu000082', 'lowest_ID']`, and `r paste(gsub("_unclassified", "", common_vars_summary['Otu000006', 'lowest_ID']))`.  The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

@@

***Treatment and Time Differences***
After observing these changes in positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).  Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible confounder.  Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).

@@

## Discussion
From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for indivdiuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, oberseved OTUs, or evenness regardless of whether the indidivual had an adenoma or carcinoma [Table S1].  There was alos no differences in relative abundnace of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, or any addition of new or depressed bacterium.  Instead it may be due to a shift of the bacterial community as a whole between existing members of the host's microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being a before sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.  There was also no difference in the thetayc distance metric based on length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.

>>>>

**Results:** The bacterial microbiome in the pre and post surgery samples of the individuals with adenoma were similar while those with carcinoma were not (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`).  There was no difference in the relative abundance of any OTU between the pre and post surgery samples (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion (AUC = `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)`) while a model built to classify samples as before surgery had `r length(IF_top10_varsMDA$variable)` variables (AUC = `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)`).  The post surgery sample for both models had a decrease in the positive probability for either lesion or before surgery sample (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and the majority of these classified to commensal bacteria (`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`).   
**Conclusions:** Our data supports the hypothesis that there are differences in the bacterial microbiome between pre and post surgical samples.  With individuals with carcinoma having more drastic changes to the overall community then those with adenoma.  However, changes to commensal bacteria were common to individuals with either adenoma or carcinoma suggesting that these bacteria may be central to initial polyp formation and transition to carcinoma.

@@

***Common OTUs to both Models***
The reduced models were built based on the most important variables to either classification model.  For the lesion model there were a total of `r length(lesion_top10_varsMDA$variable)` variables [Figure S2] whereas for the initial sample model there were a total of `r length(IF_top10_varsMDA$variable)` variables [Figure S3].  For both models FIT resulted in the largest decrease in MDA [Figure S2a & S3a].  There were a total of `r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`, `r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`, and `r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`.  The majority of these OTUs had classifications to bacteria typically thought of as commensal [Table S2].

@@

***Treatment and Time Differences***
After observing these changes in positive probability we wanted to assess whether additional treatments, such as chemotherapy and radiation, could have an impact on the results that we observed.  There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the before sample model was similar there was a significant difference in decreased positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`).  Finally, we wanted to know if the length of time between the initial and follow up sample could be a possible con-founder.  Within our study there was a significant difference in the time elapsed in the collection of the follow up sample between adenoma and carcinoma samples (uncorrected P-value = `r format(time_pvalues["time", "uncorrected_p_value"], scientific = TRUE, digits = 3)`) with time passed being lower for adenoma (`r format(time_summary['adenoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['adenoma', 'sd_days'], digits = 3)` days) then carcinoma (`r format(time_summary['carcinoma', 'mean_days'], digits = 3)` +/- `r format(time_summary['carcinoma', 'sd_days'], digits = 3)` days).

@@

## Discussion
From our results there were some large observed differences in the bacterial microbiome between pre and post surgery samples based on whether the individual had an adenoma or carcinoma.  There was much larger difference between initial and follow up samples based on the thetayc distance metric and in fecal blood as measured by FIT for individuals with carcinoma versus adenoma [Figure 1].  However, there were no differences between initial and follow up samples for Shannon Diversity, observed OTUs, or evenness regardless of whether the individual had an adenoma or carcinoma [Table S1].  There was also no differences in relative abundance of any specific OTU for lesion, adenoma, or carcinoma [Figure s1].

Although there were no differences when investigating all OTUs, when looking specifically at four OTUs that taxonomically classified to previously suggested cancer causing microbes we found that only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had a decrease in relative abundance between initial and follow up for those with carcinoma and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))`/`r length(rownames(main_crc_bugs_summary))` had differences for those with adenoma.  This data would suggest that these specific OTUs may be important in the transition of an adenoma to a carcinoma but less so in the initiation of an adenoma from benign tissue.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial samples in order to find common OTUs in the community that change for both adenoma and carcinoma.  What we found was that the commonly associated CRC bacteria were not higly represented within our models but rather that OTUs that made up the most important variables overwhelmingly belonged to commensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of commensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

*Problematic paragraph need to amend*

There was no difference for the majority of models tested for differences in positive probability based on whether chemotherapy or radiation was received.  There was a difference in the length of time between initial and follow up sample between adenoma and carcinoma.  These results would indicate that the findings described were specific to the surgical intervention and that differences observed between carcinoma and adenoma samples can not be simply attributed to collection time between samples.
